\documentclass[12pt,a4paper]{article}
\usepackage{setspace} \onehalfspacing
\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}{Proposition}
\renewcommand{\theproposition}{\arabic{proposition}}
\newtheorem{corollary}[theorem]{Corollary}
\newcommand{\real}{\mathbb{R}_+^n}

\begin{document}

\section{The model}

Two states of the world: policies $x$ and $y$. Party $A$ and $B$ prefer $x$ and $y$ respectively. We can make this more explicit with
\begin{align*}
    U_A(x) > U_A(y)\\
    U_B(y) > U_B(x)
\end{align*}
For voters, $v_i > 0$ means voter $i$ prefers $x$ to $y$ All $i \in N$ vote for $x$ or $y$. Simple majority rule determines which policy gets implemented. For each voter, each party sets a bribe schedule
\begin{align*}
    a \in (a_1, ..., a_n) \in \mathbb{R}^n_+\\
    b \in (b_1, ..., b_n) \in \mathbb{R}^n_+
\end{align*}
Solving through backward induction, given bribe schedules $(a,b)$, voter $i$ prefers to vote for $x$ if $a_i + v_i > b_i$ and for $y$ otherwise. Since indifferent voters choose $y$, party $B$ needs to only match bribes from $A$, adjusting for individual voters' preferences: $b_i = a_i + v_i$. Therefore, $B$ solves

$$\min_C \left\{\sum_{i \in C} \max\{0, a_i + v_i\} : |C| > \frac{n}{2} \right\}$$

As long as this sum is strictly less than $W_B$; otherwise party $B$ chooses to set $b_i = 0 : \forall i \in N$.

Following Banks (2000), we restrict our analysis to the set of equilibria in which party $A$ wins, i.e. $W_A$ is sufficiently large relative to $\mathbf{v}$ and $W_B$ so that policy $x$ prevails over $y$. In other words, the folloing inequality must hold:

$$ \sum_{i \in C} \max\{0, a_i + v_i\} \geq W_B$$

Let $U(v, W_b) \subseteq \mathbb{R}^n_+$ denote the set of unbeatable bribe schedules. Additionally, let $S(a) = \sum^n_i a_i$ denote the bribe schedule for party $A$. The above assumptions on $W_A$, $W_B$ and $v$ guarantee that there is an

$$\tilde{a} \in U(\mathbf{v}, W_B) : S(\tilde{a}) \leq W_A$$

For party $A$, the solution is

\begin{equation}
\label{eqn:solution}
\min_a\{S(a) : a \in U(\mathbf{v}, W_B) \}
\end{equation}

To fully describe the solution to equation \ref{eqn:solution}, we note the following: for any $a \in \mathbb{R}_+^n$, let $C(a) : {i \in N : a_i > 0}$ denote the set of individuals who receive a bribe from $A$. One can show that there is a bribe schedule $a'$ such that for any $i,j \in C(a)$, $a'_i + v_i = a'_j + v_j$. The intuition is that $A$ has no incentive to make voters differentially bribed, because $B$ will simply ignore the more expensive voters and target the weakest rings in the chain. Following Groseclose and Snyder (1996) we refer to this as a leveling schedule.

Let $U^l(\mathbf{v}, W_B) \subseteq U(\mathbf{v}, W_B)$ denote the set of unbeatable leveling schedules. These are bribe schedules such that $a_i + v_i = a_j + v_j  \equiv t(a)$. The bribe $a_i = t(a) - v_i$ is the sum of two terms. The first is the common "transfer" among all voters in $C(a)$, the second ($-v_i$) is individual specific. The latter term makes voters indifferent between $x$ and $y$ absent any bribe from $B$; the former represents the per capita amount necessary to make $C(a)$, together with any unbribed voters, unaffordable for $B$.

To further simplify the analysis, Banks introduces the folowing sets of assumption:

\begin{align*}
    A_1 &: v_{(n+1)/2} < 0\\
    A_2 &: v_1 < 2W_B/(n+1) 
\end{align*}

$A_1$ implies that absent any bribes by $A$, $y$ will defeat $x$. Therefore $A$ must bribe at least one voter. $A_2$ further implies that $A$ must bribe at least a majority of voters, otherwise $B$ will have sufficient resources to bribe $(n+1)/2$ voters and win.

Banks then proceeds to show that there are monotonic bribing schedules contained within the solution for equation \ref{eqn:solution}. For any $a \in \real$ let $k(a) = \lvert C(a) \rvert$. Suppose that $a\in U^l(\mathbf{v}, W_B)$ is such that $v_i \geq v_j$ and $j \in C(a)$ but $i \notin C(a)$. Then, under $A_2$, there exists $a' \in U^l(\mathbf{v}, W_B)$ with $S(a') \leq S(a)$, $k(a') = k(a)$ and $i \in C(a')$ but $j \notin C(a')$ by simply swapping $i$ for $j$.\footnote{Note that since $v_i \geq v_j$, we have that $t(a) - v_i \leq  t(a) - v_j$, i.e. $a'_i \leq a_j$. $A_2$ guarantees that $a'_i$ and $a_j$ are non-negative.}

Generalizing, and recalling that $v_1 \geq ... \geq v_n$, we see that for all $a \in U^l(\mathbf{v}, W_b)$ there exists a bribe schedule $a' \in U^l(\mathbf{v}, W_b)$ such that $S(a') \leq S(a)$ and $C(a') = \{1, ..., k(a)\}$. Therefore, we can without loss of generality restrict attention to schedules $a$ by $A$ which bribe the first $k(a)$ voters. Call these monotonic leveling schedules and let $U_m^l \subseteq U(\mathbf{v}, W_B)$.

Therefore, when $A_2$ holds,

$$\min\{S(a): a \in U(\mathbf{v}, W_B)\} = \min\{S(a): a \in U^l_m(\mathbf{v}, W_B)\}$$

We can now further simplify the total expenditure $S(a)$

\begin{equation*}
    S(a) = \sum_{i \in C(a)}a_i = k(a) \cdot t(a) - \sum_{i \leq k(a)}v_i
\end{equation*}

Note that the choice of $k(a)$ and $t(a)$ fully characterize any schedule $a \in U_m^l(\mathbf{v}, W_B)$. We can thus fully characterize the optimization problem of $A$ as

\begin{equation*}
    \min_{k,t} k \cdot t- \sum_{i \leq k} v_i
\end{equation*}

subject to the constraint that the induced schedule $a \in U_m^l$. Banks then reformulates this as an unconstrained problem by noting the following. First, if $a(k, t, \mathbf{v})$ is unbeatable, it must be that $k \geq (n + 1)/2$, so by $A_1$ it must be that if $a_i(k, t, \mathbf{v}) = 0$, then $v_i < 0$. Therefore, $B$ receives all non-bribed voters for free. For $a(k, t, \mathbf{v})$ to be unbeatable, then, it must be that $B$ cannot afford the remaining $(n + 1)/2 - (n - k) = k - (n - 1)/2$ voters, or

$$t * (k - (n - 1)/2) \geq W_B$$

Solving this for equality yields the optimal transfer from $A$ to members of $C(A) = \{1, ..., k\}$, conditional on k:

\begin{align}
    \label{eqn:common_transfer}
    t(k, W_B) = \frac{W_B}{k - (n - 1)/2}
\end{align}

Defining minimal winning expenditures as

\begin{align}
    \label{eqn:total_expenditure}
    E(k, \mathbf{v}, W_B) = k \cdot t(k, W_B) - \sum_{i \leq k} v_i
\end{align}

we can state $A$'s problem as

\begin{align}
\label{eqn:minimal_expenditure}
\min_k \{E(k, \mathbf{v}, W_B) : k \in {(n + 1/2), ..., n}\}
\end{align}

Denote the solution to expression \ref{eqn:minimal_expenditure} as $k^*(\mathbf{v}, W_B)$. This solution genenerates a solution to expression \ref{eqn:solution}

\section{Results}

First, characterize a solution for $k^*$. Because $k$ is finite, calculus cannot be employed. Instead, we deploy a discrete version of these techniques. First let's define $\Delta(k) = E(k + 1) - E(k)$ as the difference in expenditure from adding another coalition member. Note that if $\Delta(k) \geq 0$ then $A$ does not want to add another member to the coalition. Conversely, if $\Delta(k) < 0$, then $A$ is strictly better off by adding the $k + 1$th member of the coalition.

Next, suppose that $\Delta(k)$ is increasing in $k$. The following algorithm can then be used to identify $k^*$: if $\Delta((n + 1)/2) \geq 0$, then we know from $]\Delta(k)$ increasing that $A$ is better off by setting $k^*$ to $(n + 1)/2$. If $\Delta((n + 1)/2) < 0$, then we know that $k^*$ must be greater than $(n + 1)/2$, so we next solve for $\Delta((n + 3)/2)$, and so on.

We can therefore search for the optimal $k^*$ with the following algorithm:

\begin{align}
    \label{eqn:optimal_k}
    k^* = 
    \begin{cases}
        (n + 1)/2 & \text{if } \Delta((n + 1)/2) \geq 0 \\
        \max\{k : \Delta(k - 1) < 0\} & \text{otherwise}
    \end{cases}
\end{align}

Banks also shows that $\Delta(k)$ is indeed nondecreasing.

\begin{align}
    \Delta(k)  & = \left[\frac{(k + 1)W_B}{k + 1 - (n - 1)/2} - \sum_{i \leq k + 1}v_i \right]\\
    \label{eqn:delta_k}
    & = \frac{-W_B (n - 1)}{2(k + 1 - (n - 1)/2)(k - (n - 1)/2)} - v_{k + 1}\\
    \label{eqn:def_delta_k}
    & \equiv T(k, W_B) - v_{k + 1}
\end{align}

Note that because $v_1 \geq ... \geq v_n$, $-v_{k + 1}$ is non-decreasing in $k$. Also, taking the first term as a continuous variable and differentiating w.r.t. $k$, we note that it is also increasing in $k$.

Generating an explicit characterization of $k$ is not straightforward. Instead, Banks opts to show the conditions which need to hold in order for the coalition to be universal or minimal. Using equation \ref{eqn:optimal_k} and substituting in equation \ref{eqn:delta_k} we have the following.

\begin{proposition}
    (a) $k^*(\mathbf{v}, W_B) = (n + 1)/2$ if and only if $v_{(n + 3)/2} \leq -W_B(n - 1)/4$; (b) $k^*(\mathbf{v}, W_B) = n$ if and only if $v_n > -2W_B/(n + 1)$.
\end{proposition}

Banks also identifies how the optimal coalition $k^*$ respond to marginal changes in voter prefence intensity. Given an arbitrary amount $W_B$ and preference profile $\mathbf{v}'$, let $k' = k^*(\mathbf{v}, W_B)$. If $k' = (n + 1)/2$, then we know that $k' \leq k*(\mathbf{v}, W_B)$ for all $\mathbf{v}$, so suppose $k' > (n + 1)/2$.

From equation \ref{eqn:optimal_k} we infer that $\Delta(k - 1, \mathbf{v}', W_B) < 0)$, which from equations \ref{eqn:delta_k} and \ref{eqn:def_delta_k} is equivalent to $v_k' > T(k' - 1, W_B)$. Now suppose that the preference profile changes from $\mathbf{v}'$ to $\mathbf{v}$, and $v_{k'}$ is such that $v_{k'} \geq v'_k$. Then, $v_{k'} > T(k' - 1, W_B)$, and hence $\Delta(k' - 1, \mathbf{v}, W_B) < 0$. But then from equation \ref{eqn:optimal_k} it must be the case that $k^*(\mathbf{v}, W_B) \geq k'$. Therefore, the following holds:

\begin{proposition}
    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k'} \geq v'_{k'}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \geq k^*(\mathbf{v}', W_B)$
\end{proposition}

In words, if the preference intensity of the marginal bribed voter weakly increases, then the optimal coalition size also weakly increases. Substantively, the number of voters bribed by $A$ weakly increases as the voter who receives the largest bribe finds $A$'s preferred alternative, $x$, more attractive. Similarly

\begin{proposition}
    For all $W_B$, if $\mathbf{v}$ and $\mathbf{v}'$ are such that $v_{k' + 1} \leq v'_{k' + 1}$, where $k' = k^*(\mathbf{v}', W_B)$, then $k^*(\mathbf{v}, W_B) \leq k^*(\mathbf{v}', W_B)$
\end{proposition}

\end{document}